---
title: "Birth-Death-Immigration with gamma noise observations"
author: "André F. B. Menezes"
format: pdf
---

```{r setup, include=FALSE}
rm(list = ls())
library(ggplot2)
library(cowplot)
library(patchwork)
library(parallel)

# Source some functions and the C++ class
source("../utils.R")
source("./particleMCMC.R")
Rcpp::sourceCpp(file = "bdi.cpp")

# Knitr setup
knitr::opts_chunk$set(
    tidy = FALSE,
    comment = "#",
    collapse = TRUE, ## colapsa chunks em R Markdown
    fig.width = 12,
    fig.height = 8,
    fig.align = "center",
    fig.pos = "H",
    out.width = "100%",
    warning = FALSE,
    message = FALSE,
    echo = FALSE,
    dpi = 200, ## aumentar o dpi para exibir maior.
    cache.path = "cache/",
    fig.path = "figures/")

# Plot theme
theme_set(
  theme_cowplot(font_size = 20, font_family = "Palatino") +
    background_grid() +
    theme(text = element_text(family = "Palatino", size = 20),
          legend.position = "top")
)
```


# Birth-Death-Immigration (BDI)

The Birth-Death-Immigration (BDI) is model to describe the population of
individuals (for example, country, bacteria, etc), considering three possible 
types of events: birth, death and immigration. The model can be written using
the notation of coupled chemical reactions, as follows
\begin{align*}
\mathcal{X} &\xrightarrow{\mu} 2\mathcal{X} \quad (\text{birth})\\
\mathcal{X} &\xrightarrow{\lambda} \emptyset \quad (\text{death}) \\
\emptyset &\xrightarrow{\gamma} \mathcal{X} \quad (\text{immigration})
\end{align*}
where $\mathcal{X} = \mathbb{Z}_{+}$ denotes the state space of the process.
The parameters of the model are $\theta = (\mu,\lambda, \gamma)$. In this model,
individuals reproduce with constant hazard $\mu$, each individual dies with constant
hazard $\lambda$, and individuals arrive into the population with constant
hazard $gamma$. Consequently, the population increases by one when birth or immigration
event occurs, and decreases by one when death event occurs.

The hazard function associated to the BDI model is
$$
h(x, \theta) := \left( x\mu, x\lambda, \gamma \right).
$$


Let $X_t$ denote the number of individuals at time $t$. 
The process $\{X_t\}_{t>0}$ is naturraly describe by a Markov jump process, that is,
a continous time discrete state Markov process.
Let $dt$ denote a very small time period, the transition kernel (probabilities)
of BDI process can be written as follows:
\begin{align*}
\mathbb{P}\left\lbrack X_{t+dt} = x+1 \mid X_{t}= x\right\rbrack &=
(\gamma + x\mu)dt + o(dt)\\
\mathbb{P}\left\lbrack X_{t+dt} = x-1 \mid X_{t}= x\right\rbrack &= x\lambda dt + o(dt)\\
\mathbb{P}\left\lbrack X_{t+dt} = x \mid X_{t}= x\right\rbrack &= 1 - (\gamma + x\mu - x\lambda)dt+o(dt)\\
\mathbb{P}\left\lbrack X_{t+dt} = y \mid X_{t} = x\right\rbrack &= o(dt), \quad y\notin \{x-1, x, x+1\}
\end{align*}
where $o(dt)/dt \rightarrow 0$ as $dt \rightarrow 0$.
The instantaneous rate, or hazard function, is defined by:
$$
h(x_t, \theta) = \lim_{dt\rightarrow 0}\frac{1}{dt}\mathbb{P}\left\lbrack X_{t+dt} = x_{t+dt} \mid X_t= x_t \right\rbrack = \left( x_t\mu, x_t\lambda, \gamma \right)
$$



we can use a Markov jump
process (MJP) to describe the evolution of the state via:
$$
X_t = x_0 + \sum_{i=1}^3 S_i\,R_{i,t}
$$
where $S = (1, -1, 1)$, and $R_{i,t}$.


In a short interval $d_t$, the expected value of $X_t$ is
$$
E(dX_t) = \left(\gamma + (\mu - \lambda) x \right)dt
$$

# Simulate data


```{r sim-data}
# Settings
tmax <- 30L
dt <- 1 / 30 # Use 1/dt observations for each time period
x0 <- 500L
true_theta <- c(0.5, 0.8, 40)
hazard_bdi <- function(x, theta) c(x * theta[1], x * theta[2], theta[3])
S_bdi <- matrix(c(1, -1, 1), nrow = 1)

# Simulate
sim_xt <- sim_poisson_leap(tmax = tmax, x0 = x0, theta = true_theta,
                           hazard = hazard_bdi, S = S_bdi, dt = dt)

# Adding a observational layer
xt <- sim_xt$x[, 1L]
t_continuous <- sim_xt$tt
n <- length(xt)
retain <- c(1, seq.int(11, length(t_continuous), length.out = tmax - 1))
t_retain <- t_continuous[retain]
xt_retain <- xt[retain]

par(mfrow = c(3, 1))
true_alphas <- c(5.0, 10, 50)
y <- matrix(0.0, nrow = tmax, ncol = length(true_alphas))
for (k in seq_along(true_alphas)) {
  y[, k] <- stats::rgamma(n = tmax, shape = true_alphas[k], rate = true_alphas[k] / xt_retain)
  plot(0:29, y[, k], col = "red", ylim = c(0, max(max(y[, k]), max(xt))), pch = 19,
       ylab = "", xlab = "t", main = paste0("alpha = ", true_alphas[k]))
  lines(t_continuous, xt)
  lines(0:29, xt_retain, col = "blue", pch = 18)
}
```


# Inference only for $\mu$

```{r ff, results='hide', cache=TRUE}
NDPOST <- 10000L
NSKIP <- 10000L
N_PARTICLES <- 40L
MAX_TRIALS <- 300L
TARGET_SUCCESS <- 40L
na <- length(true_alphas)
tab_bf <- tab_ff <- matrix(nrow = na, ncol = 6L)
draws_bf <- matrix(0.0, nrow = NDPOST, ncol = na)
draws_ff <- matrix(0.0, nrow = NDPOST, ncol = na)
for (k in seq_len(na)) {
  res_bf <- particleMCMC_bdi(y = y[, k], x0 = x0, dt = 1/10,
                             filter = "bootstrap",
                             n_particles = N_PARTICLES,
                             shape = true_alphas[k],
                             ndpost = NDPOST, nskip = NSKIP,
                             theta_fix = true_theta[-1L],
                             wh_fix = c(2L, 3L),
                             sd_prior = 1.0,
                             S_prop = 0.5,
                             printevery = 100L, theta_init = true_theta[1L])
  draws_bf[, k] <- res_bf[, 1L]
  # Run FF
  res_ff <- particleMCMC_bdi(y = y[, k], x0 = x0, dt = 1/10,
                          filter = "franken",
                          target_success = TARGET_SUCCESS,
                          max_trials = MAX_TRIALS,
                          shape = true_alphas[k],
                          ndpost = NDPOST, nskip = NSKIP,
                          theta_fix = true_theta[-1L],
                          wh_fix = c(2L, 3L),
                          sd_prior = 1.0,
                          S_prop = 0.5,
                          printevery = 10L, theta_init = true_theta[1L])
  draws_ff[, k] <- res_ff[, 1L]
  # Compute ESS
  ess_bf <- coda::effectiveSize(coda::as.mcmc(res_bf))
  ess_ff <- coda::effectiveSize(coda::as.mcmc(res_ff))
  # Save info
  tab_bf[k, 1L] <- ess_bf
  tab_bf[k, 2L] <- attr(res_bf, "time")[3L]
  tab_bf[k, 3L] <- ess_bf/attr(res_bf, "time")[3L]
  tab_bf[k, 4L] <- mean(res_bf)
  tab_bf[k, 5L] <- quantile(res_bf, probs = 0.025)
  tab_bf[k, 6L] <- quantile(res_bf, probs = 0.975)
  # FF
  tab_ff[k, 1L] <- ess_ff
  tab_ff[k, 2L] <- attr(res_ff, "time")[3L]
  tab_ff[k, 3L] <- ess_ff/attr(res_ff, "time")[3L]
  tab_ff[k, 4L] <- mean(res_ff)
  tab_ff[k, 5L] <- quantile(res_ff, probs = 0.025)
  tab_ff[k, 6L] <- quantile(res_ff, probs = 0.975)
}

```

```{r inference, results='hide', cache=TRUE}
# NDPOST <- 8000L
# NSKIP <- 6000L
# N_PARTICLES <- 40L
# MAX_TRIALS <- 300L
# TARGET_SUCCESS <- 40L
# na <- length(true_alphas)
# tab_bf <- tab_ff <- matrix(nrow = na, ncol = 6L)
# draws_bf <- matrix(0.0, nrow = NDPOST, ncol = na)
# draws_ff <- matrix(0.0, nrow = NDPOST, ncol = na)
# for (k in seq_len(na)) {
#   res_bf <- particleMCMC_bdi(y = y[, k], x0 = x0, dt = 1/10,
#                              filter = "bootstrap",
#                              n_particles = N_PARTICLES,
#                              shape = true_alphas[k],
#                              ndpost = NDPOST, nskip = NSKIP,
#                              theta_fix = true_theta[-1L],
#                              wh_fix = c(2L, 3L),
#                              sd_prior = 1.0,
#                              S_prop = 0.5,
#                              printevery = 100L, theta_init = true_theta[1L])
#   draws_bf[, k] <- res_bf[, 1L]
#   # Run FF
#   res_ff <- particleMCMC_bdi(y = y[, k], x0 = x0, dt = 1/10,
#                           filter = "franken",
#                           target_success = TARGET_SUCCESS,
#                           max_trials = MAX_TRIALS,
#                           shape = true_alphas[k],
#                           ndpost = NDPOST, nskip = NSKIP,
#                           theta_fix = true_theta[-1L],
#                           wh_fix = c(2L, 3L),
#                           sd_prior = 1.0,
#                           S_prop = 0.5,
#                           printevery = 10L, theta_init = true_theta[1L])
#   draws_ff[, k] <- res_ff[, 1L]
#   # Compute ESS
#   ess_bf <- coda::effectiveSize(coda::as.mcmc(res_bf))
#   ess_ff <- coda::effectiveSize(coda::as.mcmc(res_ff))
#   # Save info
#   tab_bf[k, 1L] <- ess_bf
#   tab_bf[k, 2L] <- attr(res_bf, "time")[3L]
#   tab_bf[k, 3L] <- ess_bf/attr(res_bf, "time")[3L]
#   tab_bf[k, 4L] <- mean(res_bf)
#   tab_bf[k, 5L] <- quantile(res_bf, probs = 0.025)
#   tab_bf[k, 6L] <- quantile(res_bf, probs = 0.975)
#   # FF
#   tab_ff[k, 1L] <- ess_ff
#   tab_ff[k, 2L] <- attr(res_ff, "time")[3L]
#   tab_ff[k, 3L] <- ess_ff/attr(res_ff, "time")[3L]
#   tab_ff[k, 4L] <- mean(res_ff)
#   tab_ff[k, 5L] <- quantile(res_ff, probs = 0.025)
#   tab_ff[k, 6L] <- quantile(res_ff, probs = 0.975)
# }
```

```{r tab}
colnames(tab_bf) <- colnames(tab_ff) <- c("ess", "sec", "ess/s", "mean", "lo", "up")
tab_bf <- as.data.frame(tab_bf)
tab_ff <- as.data.frame(tab_ff)
tab_bf$pf <- "BF"
tab_ff$pf <- "FF"
tab_ff$shape <- true_alphas
tab_bf$shape <- true_alphas
tab <- rbind(tab_bf, tab_ff)
tab <- tab[, c("shape", "pf", "ess", "sec", "ess/s", "mean", "lo", "up")]
tab
```


# Parameter inference for $\mu$ and $\gamma$
















